---
title: "final project"
output:
  pdf_document: default
  html_document: default
date: "2023-06-20"
---

# Load the packages and data

```{r setup libraries, include=FALSE}

library(dplyr)
library(ggplot2)
library(modelr)
library(lubridate)
library(tidyr)
```


```{r setup data, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
city_day_agg=read.csv(gzfile("city_day_agg_cleaned.csv.gz"))
country_day_agg=read.csv(gzfile("country_day_agg_cleaned.csv.gz"))

head(city_day_agg)
head(country_day_agg)
```

## play with USA data
```{r usa cities}
city_day_agg%>%
  filter(countryCode=="USA",parameter=="pm25")%>%
  mutate(date=as.Date(date))%>%
  count(date)%>%
  ggplot(aes(x=date,y=n))+
  geom_point()
```

certain days in usa have more data points collected
```{r usa from country}
#takeaway: there are null values in the dataset
country_day_agg%>%
  filter(countryCode=="USA")%>%
  ggplot(aes(x=as.Date(date),y=mean,color=parameter))+
  geom_point()+
  geom_smooth()+
  scale_y_log10()+
  xlab("Date")+
  ylab("values")
#missing values for some days, see geom_smooth parts

country_day_agg%>%
  filter(countryCode=="USA")%>%
  mutate(date=as.Date(date),yr_mo=format(as.Date(date), "%Y-%m"))%>%
  group_by(parameter,yr_mo)%>%
  summarize(count=n(),values=mean(mean))%>%
  ggplot(aes(x=yr_mo,y=values,color=parameter))+
  geom_point()+
  geom_smooth()+
  scale_y_log10()+
  xlab("Date")+
  ylab("values")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
separare the data into prev covid and during covid

```{r usa from country+covid19 factor}
country_day_agg%>%
  filter(countryCode=="USA")%>%
  mutate(during_covid=date>"2019-12-31")%>%
  group_by(during_covid,parameter)%>%
  summarize(min(mean),max(mean))


country_day_agg%>%
  filter(countryCode=="USA")%>%
  mutate(during_covid=date>"2019-12-31")%>%
  ggplot(aes(x=mean,color=during_covid,fill=during_covid))+
  geom_density(position = "identity",alpha=0.5)+
  facet_wrap(~parameter)+
  scale_x_log10()
```

for usa,we can see that 
(1) in 2020, NO2 level decreases b/c we see more days with lower NO2 levels;
(2) 03 level increases in 2020 b/c the distribution is about the same for prev-covid and 2020 but we see more days with high 03 values;
(3)pm2.5 level decreases in 2020 b/c we see more days with less pm2.5 levels



# play with the world dataset

here is the one for the country dataset
## country datasets
```{r world, echo=FALSE}
median_values=country_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  group_by(countryCode,in2020,parameter)%>%
  summarize(avg_value=mean(mean))%>%
  ungroup()%>%
  group_by(in2020,parameter)%>%
  summarize(med_val=median(avg_value,na.rm = TRUE))


#we only wants months from 1-5
country_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  group_by(countryCode,in2020,parameter)%>%
  summarize(avg_value=mean(mean))%>%
  left_join(median_values,by=c("in2020","parameter"))%>%
  ggplot(aes(x=avg_value,color=in2020,fill=in2020))+
  geom_density(position = "identity",alpha=0.5)+
  facet_wrap(~parameter)
```

even though our graph looks different from what authors have on the papers, the trends are the same:
1. for no2, in2020 graph shifts to the left, which means no2 levels decreases overall.
2. for o3, in2020 graph shifts to the right, which means 03 level increases overall
3. for pm25, in2020 graph shifts to the left, which means pm25 level decreases overall.
```{r world, echo=FALSE}

#add medians to the graph
add_medians=function(par){
  df=country_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  filter(parameter==par)%>%
  group_by(countryCode,in2020)%>%
  summarize(avg_value=mean(mean))
  
  median=df%>%
    ungroup()%>%
    group_by(in2020)%>%
    summarize(med=median(avg_value,na.rm=TRUE))
  
  df%>%
    ggplot(aes(x=avg_value,color=in2020,fill=in2020))+
    geom_density(position = "identity",alpha=0.5)+
    geom_vline(data=median,aes(xintercept=med,color=in2020))
}

add_medians("pm25")

```
unequal number of countries before 2020 and after 2020

```{r world+rows check, echo=FALSE}
country_day_agg%>%
  filter(date>"2019-12-31")%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  select(countryCode)%>%
  unique()%>%
  nrow()

country_day_agg%>%
  filter(date<"2019-12-31")%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
   filter(jan_may==TRUE)%>%
  select(countryCode)%>%
  unique()%>%
  nrow()
```
## here is the graph using city dataset

overall patterns in no2, o3, and pm25 are less obvious
density distribution is very clustered

```{r cities dataset, echo=FALSE}
city_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  group_by(city_id,in2020,parameter)%>%
  summarize(mean=mean(mean))%>%
  ggplot(aes(x=mean,color=in2020,fill=in2020))+
  geom_density(position = "identity",alpha=0.5)+
  facet_wrap(~parameter)
```
separate data into two dataframes, one before 2020 and one after 2020

```{r world country check, echo=FALSE}
#during pandemic, we have 50 countries
cities_2020=city_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  filter(in2020==TRUE)%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  #count(country)%>%
  #nrow()
  group_by(countryCode,city_id,parameter)%>%
  summarize(yr2020=mean(mean))

#before pandemic, we have 56 countries 
cities_yr=city_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  filter(in2020==FALSE)%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  #count(country)%>%
  #nrow()
  group_by(countryCode,city_id,parameter)%>%
  summarize(yr3_avg=mean(mean))
```


```{r world country check, echo=FALSE}
#filter out countries that are common in 20202 and before 2020
inner_join(cities_2020,cities_yr,by=c("countryCode","city_id","parameter"))%>%
  #filter(parameter=="pm25")%>%
  group_by(countryCode,city_id,parameter)%>%
  ggplot()+
  geom_density(aes(x=yr2020,color="pink",fill="pink",alpha=0.5))+
  geom_density(aes(x=yr3_avg,color="turquoise",fill="turquoise",alpha=0.5))+
  #geom_vline(aes(xintercept=mean(cities_yr$yr3_avg)))+
  #geom_vline(aes(xintercept=median(cities_2020$yr2020)))
  facet_wrap(~parameter)


```


```{r world, echo=FALSE}
add_medians2=function(par){
  

  df=city_day_agg%>%
  mutate(in2020=date>"2019-12-31")%>%
  mutate(date=as.Date(date))%>%
  mutate(jan_may=month(date)<=5)%>%
  filter(jan_may==TRUE)%>%
  filter(parameter==par)%>%
  group_by(city_id,in2020)%>%
  summarize(avg_value=mean(mean))

  
  median=df%>%
    ungroup()%>%
    group_by(in2020)%>%
    summarize(med=median(avg_value,na.rm=TRUE))
  
  df%>%
    ggplot(aes(x=avg_value,color=in2020,fill=in2020))+
    geom_density(position = "identity",alpha=0.5)+
    geom_vline(data=median,aes(xintercept=med,color=in2020))
}
add_medians2("pm25")
  
```
after i aggregate city data to get country data, the pattern becomes less obvious.



# mapping

```{r world map, echo=FALSE}
WorldData <- map_data('world') 

lon_lat_data=read.csv("openaq_cities.csv")

df_datasets=inner_join(cities_2020,cities_yr,by=c("countryCode","city_id","parameter"))%>%
  left_join(lon_lat_data,by=c("countryCode","city_id"))%>%
  mutate(perc_chg=(yr2020-yr3_avg)/yr3_avg*100)
  
  
per_chg_plot=function(par){
df_t=df_datasets%>%filter(parameter==par)
ggplot()+
  geom_polygon(data=WorldData,aes(x=long,y=lat,group=group),fill = "white", colour = "grey50")+
  geom_point(data=df_t,aes(x=Lon,y=Lat,color=perc_chg),shape=1)+
  coord_cartesian(xlim = c(-115,145),ylim=c(-55,70))+
  scale_color_continuous(high="orange",low="purple",limits=c(-20,20))
}
per_chg_plot("pm25")
```
my graph has more points compared with maps on Fig1. We have more cities.

# modeling
```{r modeling usa, echo=FALSE}
df_in2020_ornot=country_day_agg%>%
  mutate(in2020=date>"2019-12-31",mon=month(as.Date(date),label=TRUE),wkd=weekdays(as.Date(date)))%>%
  mutate(is_weekend=wkd %in% c("Saturday","Sunday"))

#usa data in2020
df2020_usa=df_in2020_ornot%>%
  filter(in2020==TRUE,countryCode=="USA")%>%
  filter(date>="2020-03-15")%>%
  drop_na()%>%
  group_by(parameter,mon,is_weekend)%>%
  summarize(mean=mean(mean,na.rm = TRUE))

#usa data before 2020
df_b420_usa=df_in2020_ornot%>%
  filter(in2020==FALSE,countryCode=="USA")%>%
  filter(mon %in% c("Mar","Apr","May"))%>%
  drop_na()%>%
  #group_by(parameter,year(as.Date(date)),mon,is_weekend)%>%
  group_by(parameter,mon,is_weekend)%>%
  summarize(mean=mean(mean,na.rm = TRUE))

#modeling usa
model_usa=lm(mean~parameter+mon+is_weekend,df_b420_usa)
#rmse(model_usa,df_b420_usa)
df_b420_usa%>%
  add_predictions(model_usa)%>%
  group_by(parameter)%>%
  summarize(rmse=sqrt(mean((pred-mean)^2)),
            cor=cor(pred,mean),
            cor_sq=cor^2)
```


```{r modeling switzerland, echo=FALSE}
#sw data in2020
df2020_sw=df_in2020_ornot%>%
  filter(in2020==TRUE,countryCode=="CHE")%>%
  filter(date>="2020-03-16")%>%
  drop_na()%>%
  group_by(parameter,mon,is_weekend)%>%
  summarize(mean=mean(mean,na.rm = TRUE))

#sw data before 2020
df_b420_sw=df_in2020_ornot%>%
  filter(in2020==FALSE,countryCode=="CHE")%>%
  filter(mon %in% c("Mar","Apr","May"))%>%
  drop_na()%>%
  group_by(parameter,mon,is_weekend)%>%
  summarize(mean=mean(mean,na.rm = TRUE))

#modeling usa
model_sw=lm(mean~parameter+mon+is_weekend,df_b420_sw)
df_b420_sw%>%
  add_predictions(model_sw)%>%
  group_by(parameter)%>%
  summarize(rmse=sqrt(mean((pred-mean)^2)),
            cor=cor(pred,mean),
            cor_sq=cor^2)

```


```{r modeling, echo=FALSE}
#compare modeling data with actual data
usa_chg=df2020_usa%>%
  add_predictions(model_usa)%>%
  mutate(per_chg=(mean-pred)/mean*100)%>%
  group_by(parameter)%>%
  summarize(chg=mean(per_chg),se=sd(per_chg)/sqrt(length(per_chg)))%>%
  mutate(country="USA")

sw_chg=df2020_sw%>%
  add_predictions(model_sw)%>%
  mutate(per_chg=(mean-pred)/mean*100)%>%
  group_by(parameter)%>%
  summarize(chg=mean(per_chg),se=sd(per_chg)/sqrt(length(per_chg)))%>%
  mutate(country="CHE")

rbind(usa_chg,sw_chg)%>%
  ggplot(aes(x=chg,y=country))+
  geom_point(aes(color=chg>0))+
  facet_wrap(~parameter)+
  geom_pointrange(aes(y=country,xmin=chg-se,xmax=chg+se))+
  geom_vline(xintercept = 0,linetype="dotted")+
  scale_x_continuous(limits=c(-150,100))
  
  
```

# future research
- factor in pop density data for modeling
- 
```{r modeling, echo=FALSE}



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
